---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# NPSpecies

<!-- badges: start -->
<!-- badges: end -->

NPSpecies contains "Information on Species in National Parks". The dataset was collected across all park listed here:
https://irma.nps.gov/NPSpecies/Search/SpeciesList    


## Installation

You can install the development version of NPSpecies like so:

``` r
devtools::install_github("frankiethull/NPSpecies")
```

## Example

This showcases the data format of what is typically downloaded through the NPSpecies GUI. The GUI only allows one park to be downloaded at a time. Here, we have all data for all parks. Allowing for quite a bit of exploratory data analysis across NPS. 

filtering for the most observed reptiles at Great Sand Dunes National Park and Preserve:
```{r example1}
library(NPSpecies)
library(tinytable)

NPSpecies::species |>
    dplyr::filter(ParkCode == "GRSA") |>
    dplyr::filter(CategoryName == "Reptile") |>
    dplyr::select(ParkName, CategoryName, Family, SciName, CommonNames, Observations) |>
    dplyr::arrange(
      dplyr::desc(Observations)
      ) |>
    head(3) |>
    tt()
```

filtering for top 5 most observed non-native species at Everglades National Park:
```{r example2}
library(NPSpecies)

NPSpecies::species |>
    dplyr::filter(ParkCode == "EVER") |> 
    dplyr::filter(Nativeness == "Non-native") |>
    dplyr::select(ParkName, CategoryName, Family, SciName, CommonNames, Observations) |>
    dplyr::arrange(desc(Observations)) |>
    head(5) |>
    tt()
```


most observed bird species by park across all parks:
```{r example3}
NPSpecies::species |>
  dplyr::filter(CategoryName == "Bird") |> 
  dplyr::group_by(ParkName) |>
  dplyr::select(ParkName, CommonNames, Observations) |>
  dplyr::arrange(desc(Observations)) |>
  dplyr::slice(1) |>
  dplyr::ungroup() |>
  dplyr::arrange(desc(Observations)) |> 
  head(10) |> # only display 10
  tt()
```


##### citation:
 - *this package is static data, as of Labor Day 2024, for EDA purposes*        
NPSpecies - The National Park Service biodiversity database.   
https://irma.nps.gov/npspecies/       
Accessed September 2nd, 2024      
